Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes

### 2024-01-25 Journal-manuscript/Journal-manuscriptV0.tex
R2 Journal-manuscript revision. For AP review.

Accessibility is a key concept in the analysis of land use and
transportation systems {[}e.g., 1,2,3{]} and is coming of age from the
perspective of planning research {[}see \emph{inter alia}, 4,5--8{]}.
Beginning with the work of Hansen {[}1{]}, accessibility measures have
been widely used to evaluate the efficiency of transportation systems
when combined with the distribution of populations and opportunities in
space {[}9{]}. In this way, accessibility is a holistic measure of
spatial systems' ease of reaching population-relevant destinations
{[}10,11{]}.

The most common form of accessibility measure is based the gravity model
{[}e.g. 1{]}: these measures sum ``weighted'' opportunities around a
focal point (i.e., a potential origin), based on how expensive it is to
reach them. Recent research in accessibility analysis has explored the
implications of methods by which opportunities are summed, especially in
the context of competitive accessibility {[}15{]}. In a typical
Hansen-type accessibility measure, the sums around origins are not
constrained. This means that the same opportunity can enter the sum for
different origins; Counting the same opportunity multiple times treats
the opportunity as if it was inexhaustible (or non-rival {[}16{]}).
However, this conceptualisation may not reflect reality as
opportunities, in general, are not inexhaustible. Some opportunities are
by definition exclusive: a typical example is employment, once a job is
taken up by someone in the population, the same job is no longer
available for any other person to take {[}14,17{]}. More generally, it
could be conceptualised that all opportunities are subject to some
amount of congestion or capacity-constraint: multiple people use a
specific opportunity at a given time and the more people who do, the
more congested the opportunity. Congestion can be seen on a continuum,
and debate on how congestion within accessibility should be considered
is ongoing {[}14,17,18{]}. Nonetheless, competition has been widely
considered for traditionally exclusive opportunities types such as
employment and healthcare opportunities {[}14,17,19--21,21--24{]} and
more recently educational facilities {[}18,25{]}, and less exclusive
amenities-types such as greenspace and recreation {[}25,26{]}, and
shopping {[}18,27{]}.

The consideration of congestion in accessibility measures was the
motivation for the development of approaches that consider competition
{[}21,23,24{]} and are popularly applied as floating catchment area
methods {[}e.g. 28{]}. While these approaches purport to account for
congestion, Paez et al. {[}12{]} demonstrates that they do not solve the
issue of multiple counting of opportunities in general, thus leading to
biases in the calculation of total demand and supply. They sometimes
inflate counts and other times deflate them. In response to this, recent
research has paid closer attention to the way opportunities are counted
in accessibility analysis. Paez et al. {[}12{]}, for example, focuses on
floating catchment area methods and introduces a normalization of the
impedance matrix to allocate the population and then the level of
service proportionally. More recently, Soukhov et al. {[}15{]}
introduced a singly-constrained measure of accessibility, called spatial
availability, that employs a similar but more sophisticated proportional
allocation mechanism. The work of these authors show that floating
catchment area methods can be seen as singly-constrained accessibility
measures and improve on existing approaches by guaranteeing that each
opportunity is counted only once. In other words, spatial availability
treats the opportunities as \emph{finite}. The proportional allocation
mechanism of spatial availability constrains the calculations to match a
known quantity, therefore ensuring that the measurements sum up to a
predetermined quantity (i.e., the total number of opportunities), and so
each value can be meaningfully related to this total.

@@

This paper's primary motivation is to extend spatial availability for
the case of multimodal accessibility (i.e., travel by different modes).
It is worth noting though that the consideration for travel costs by
different modes is in fact just one case of heterogeneous populations.
The method itself can easily accommodate other forms of heterogeneity,
for example: variations in travel behavior between older and younger
adults {[}e.g., 29{]}, the propensity of older adults to use different
modes of transportation {[}e.g., 30{]}, the usually shorter trip lengths
of children compared to adults {[}e.g., 31{]}, or the more limited
travel ranges of single parents {[}e.g., 32{]}.

The rest of this paper is organized as follows. In Section 2 we provide
a brief review of multimodal accessibility. In the following Section 3
we demonstrate the derivation of the new spatial availability expression
for multiple modes. In Section 4 we illustrate relevant issues through a
synthetic example. This is followed in Section 5 by an empirical example
using home-to-work data from the city of Madrid, Spain after the
implementation of its Low Emission Zones (LEZ). Data for this example is
sourced from the city's 2018 travel survey. The empirical example
demonstrates the differences in multimodal spatial availability
highlighting differences within and outside the LEZ for travelers using
different modes, namely car, transit, cycling and walking. In Section 6,
we provide concluding remarks on the strengths of the use of spatial
availability as a multimodal accessibility measure, and discuss
potential future uses in policy planning scenarios as well as directions
for future research.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used are
measures based on the gravity model {[}33{]}; this includes cumulative
opportunity measures which are a special case of the gravity model
{[}5,10{]}. These gravity model based measures weight opportunities
depending on how easy it is to reach them. Given an origin (\(i\)) and a
destination (\(j\)), an impedance function \(f^{m}(c^m_{ij})\) converts
the cost of travel (e.g., time, money, generalized cost) into a score
that represents the propensity for potential interaction. These measures
originate from the work proposed by Hansen {[}1{]}, which can take the
following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\). In this form, \(m\) is a set of
modes that have mode-specific travel costs (\(c_{ij}^m\)) and/or travel
impedance functions (\(f^m(\cdot)\)).

Hansen-type accessibility is not constrained, which is to say it does
not consider the opportunities as finite. To cite an example, Tahmasbi
et al. {[}34{]} use Hansen-type accessibility to assess the potential
interaction with retail locations by three modes: walking, public
transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of retail
locations \(j\) that can potentially be reached under the travel
impedance as calculated for each \(i\) and \(m\). In other words, for
each origin \(i\) three accessibility scores are calculated. In this
work, Tahmasbi et al. {[}34{]} show that car travel affords the highest
\(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
car can potentially reach more retail opportunities than populations
using other modes. However, higher \(S_i^{m}\) values for car do not
affect the values of \(S_i^{m}\) for other modes: in effect, each mode
is analysed as if the others did not exist. Since the measure is not
constrained, each opportunity is typically counted multiple times within
and between modes, and as a result the sum of accessibility is not
necessarily a meaningful quantity. The accessibility scores for the
modes are often values that are difficult to interpret beyond making
statements about relative size. For example, Lunke {[}35{]}, researching
the region of Oslo, reports accessibility scores for car in the order of
tens of thousands of employment opportunities. The corresponding scores
for transit are lower, but still often in the thousands or tens of
thousands. As reported, the ratio of the transit to the car score can be
lower than 0.2 (meaning transit gives access to less than 20\% of the
opportunities than car). But despite the discussion about ``sufficient
accessibility'', it is unclear what the unconstrained scores mean: is
having access to but 10,000 jobs by transit insufficient? After all,
10,000 employment opportunities are still plenty of opportunities. These
ratios can be found elsewhere in the literature {[}e.g., 36, and
9,29,32, and 8{]}, and they are useful as relative assessment of when
some members of the public are better or worse off than others, but they
are silent on how bad is ``worse off''.

Besides ratios of accessibility, another way seen in the literature to
improve interpretability of scores is to standardise them within a
{[}0-1{]} range. This adjustment is only helpful insofar as it
facilitates relative comparisons, but interpretation of the scores
remains challenging because the values are specific to a region and
convey no meaning about the magnitude of the scores. In this approach,
zones always have values between 0 and 1, but how remarkable is a zone
with a low score for pedestrians and a high value for car? And if
remarkable, what does the difference in these standardized values mean
for planners? By how much should transport systems and land-use
configurations be changed to improve conditions? And in what way can
these scores be used to track differences over time? Or between regions?
These questions lack straightforward answers since certain values will
always be relatively `low' or `high', but do not track to a quantity
that can be intuitively understood. Presentation or discussion of
Hansen-type accessibility that has been standardised in this way is not
uncommon in the literature {[}e.g., 37,38{]}.

Once we understand opportunities to be finite, it is possible for an
accessibility measure to take on a crisper meaning. As considered in the
long tradition of accessibility research, capacity of opportunities is
limited and thus is subject to competition by population
{[}14,15,17--25,39--42{]}. There are only so many school-seats, hospital
capacity, employment opportunities, etc., in a region and if one person
reaches an opportunity, it is taken: the supply of an opportunity and
the demand for that opportunity are two components of accessibility.
These are clear examples of opportunities that are unambiguously
competitive. But we would go as far as to argue that every type of
opportunity is subject to congestion or capacity constraint, even when
the opportunities are conventionally seen as non-competitive.

Amenities are a good example of this. For instance, standards for
providing green spaces are often stated in the form of \emph{exclusive
access}, in units of amenity per capita. A case in point is a the
Ile-de-France region, a jurisdiction that suggested in a 2013 planning
document that at the municipal level at least 10 \(m^2\) of public green
space should be supplied \emph{per inhabitant} {[}43{]}. Green spaces
are not evenly distributed, which means that who has access to them
hinges on where they are and how easy is to reach them. This formulation
of provision of amenities is not unusual. For example, Natural England,
an organization that recommends an Accessible Natural Greenspace
Standard such that the minimum supply of space is one ha of statutory
Local Nature Reserves per thousand population\footnote{see
  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
Similarly, the World Health Organization {[}cited in 44{]} recommends
that cities provide a minimum of 9 \(m^2\) of green area per inhabitant.
For our purposes, standards of this type translate into ``how much of
this resource is available to one individual that has not been claimed
by anyone else?''. Green spaces often have large capacities, but they
still have a capacity, and it is not the same for a person to have
access to 5 \(m^2\) of \emph{uncongested} green space as 15 \(m^2\).
This difference is in fact a matter of justice {[}43,45{]}. Constraining
accessibility is in this way a useful way to evaluate the congested
availability of any type of opportunity. As development of sound
standards is emphasized in the planning literature, in particular in
regards to fairness in transportation {[}see 46{]}, spatial availability
analysis is a useful way to develop and assess standards.

The relevance of the considerations above is put in sharper relief when
we think about the use of multiple modes (or heterogeneous populations).
If we return to Oslo for a moment {[}35{]}, we notice that the places
that have high accessibility by transit are also the places that have
\emph{very high} accessibility by car (in their Figure 2). Those two
populations are going for the same opportunities, and those travelling
by transit have fewer to choose from the start. More generally, people
in a zone who are advantaged with relatively low cost of travel will
have the ability to potentially reach more opportunities than other
people. Due to this advantage, through the perspective of finite
opportunities, there are fewer opportunities left for everyone else,
especially for those who use modes that are slower or otherwise more
expensive.

As noted in the Introduction, competitive accessibility was the
rationale for developing floating catchment area methods (FCA),
popularized by Luo et al. {[}28{]} who reformulated the work of Shen
{[}24{]} into two steps (although similar, and earlier, developments are
found in {[}21,23{]}). Shen-type accessibility is formulated as:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
variables are repeated in the Hansen-type measure. Shen-type modal
accessibility (\(a_i^m\)) can be understood as a ratio of the travel
impedance-weighted supply of opportunities for \(m\)-mode in \(i\) over
the travel impedance-weighted demand for opportunities. In this way, it
considers competition. That said, the measure remains unconstrained,
meaning both population \emph{and} opportunities are multiply counted
{[}see 12{]}. In other words, interpretation of the Shen-type
accessibility scores between modes is fraught, as it is for Hansen-type
measures.

To illustrate, {[}47{]} calculates \(a_i^m\) to jobs for different
income-group populations in Shenzhen, China using
\(m = \text{public transit}\) and \(m=\text{car}\). Their results
indicate that zones with low-income populations have lower \(a_i^m\)
than zones with higher-income populations. Further, they show that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many zones, arguing that this may further place those zones with
lower-income populations at a disadvantage. \(a_i\) and/or \(a_i^m\) are
used to compare relative spatial differences in overall competitive
accessibility and multimodal competitive accessibility, but because
opportunities were doubly counted (entering the sums of both modes),
this makes for uneasy interpretations of the differences in \(a_i^{m}\)
between modes. Questions that this approach leaves unaddressed include:
what is the impact of competition on the difference in \(a_i^m\) values?
How does the impact vary spatially? And what is the interpretation of
this difference?

Spatial availability improves on the discussed Hansen-type and Shen-type
accessibility approaches by constraining the sum of opportunities, that
is, by treating opportunities as finite. This is done by means of
proportional allocation factors that follow well established principles
of spatial interaction and the gravity model {[}see 48{]}. In Soukhov et
al. {[}15{]} these factors consider: the mass effect (e.g., the size of
populations at different origins); and the cost of travel from different
zones (e.g., some sub-populations face relatively higher or lower
costs). The following section introduces the multimodal form of spatial
availability.

@@

Compared to Hansen-type accessibility
\(S_i = \sum_{j=1}^J O_jf(c_{ij})\), we see that spatial availability
is, like the Hansen-type measure, a weighted sum of the opportunities.
What makes spatial availability stand apart from other approaches is how
the weight used in the sum (balancing factor \(F^t_{ij}\)) implements a
proportional allocation mechanisms to ensure that the sum of \(V_i\) is
constrained to match the total number of opportunities in the region. As
such, spatial availability is singly-constrained and natively considers
competition. \(F^t_{ij}\) consists of two parts. The first part is a
population-based proportional allocation factor to model the mass effect
of the gravity model: \[
F^p_{i} = \frac{P_i}{\sum_i P_i}
\]

@@

The focus of this paper is to extend \(V_i\) for the measurement of
multimodal applications (or more generally heterogeneous populations).
To do so, the balancing factors are reformulated so that 1) the mass
effect now accounts not only for the size of the population at \(i\),
but also the size of sub-populations within \(i\); and 2) the cost of
travel is not only for different zones, but by sub-populations within
each zone (e.g., the cost of travel from \(i\) by car, transit, walking,
etc.) relative to all zones. When we introduce modes (or
sub-populations) \(m\), the proportional allocation factors need to
satisfy the condition that \(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be
summed across each \(m\) at each \(i\) and then across all \(i\) to
equal to 1. They are also similarly combined multiplicatively to obtain
their joint effect, represented as the combined balancing factor
\(F^{tm}_{ij}\) similar to that detailed in Equation
(\ref{eq:multimodal-balancing-factors}). This factor is given by:

@@

Next we use a synthetic example to contrast multimodal spatial
availability with multimodal versions of Hansen-type accessibility
(unconstrained) and Shen-type (unconstrained and competitive)
accessibility.

@@

Consider the simple system shown in Figure \ref{fig:Fig1}. The figure
shows a region with population at three population centers (\(A\),
\(B\), \(C\)) and jobs at three employment centers (\(1\), \(2\),
\(3\)). The population at each origin \(i\) is consists of two
sub-populations, one using a faster mode \(z\) and another using a
slower mode \(x\), to travel to employment centers. Population center
\(A\) is Suburban: it is closest to its own relatively large employment
center at \(1\), close to the Urban's equally large employment center
\(2\), and has a population that is smaller than the Urban \(B\) and
larger than the Satellite \(C\). \(B\) has the largest \(x\)-using
population, followed by then \(A\), then \(C\). This synthetic example
is inspired by the single-mode example used in {[}24{]} and reconfigured
in {[}15{]}.

@@

From the perspective of access to a \emph{finite} amount of
opportunities in the region (\(210,000\) jobs), the sub-population that
is most proximate to jobs (low cost to reach), furthest from large
populations (less competition), and uses the fastest mode \(z\) (greater
range) can potentially reach the largest number of opportunities. This
appears to be the sub-population at \(A\) using mode \(z\).
Sub-populations located in opposite conditions (i.e., more distant from
jobs, close to large populations, and using slower mode \(x\)) are at a
relative disadvantage. The competition for opportunities between
different mode-using populations matters as it reflects how well the
land-use and transport system serves (or does not serve) certain
populations.

@@

The values calculated for \(S_i^m\) (Hansen-type accessibility),
\(a_i^m\) (Shen-type accessibility), and \(V_i^m\) (spatial
availability) for each \(i\) and \(m\) are shown in the middle three
columns and are aggregated for each \(i\) in the final two columns in
Table 1 . As in the example in Shen {[}24{]}, we use a negative
exponential impedance function
\(f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
\(x\) and \(z\) modes for all accessibility measures calculations.
Notice that in this example we use the same impedance function but the
travel times are different for the two modes. More generally, it is
possible to use different impedance functions for the modes, as
demonstrated in the empirical example in the following section.

>>>>

Accessibility is a key concept in the analysis of land use and
transportation systems {[}e.g., 1,2,3{]}, and too is coming of age from
the perspective of planning {[}see \emph{inter alia}, 4,5--8{]}.
Beginning with the work of Hansen {[}1{]}, accessibility measures have
been widely used to evaluate the efficiency of transportation systems
when combined with the distribution of opportunities in space {[}9{]}.
As such, it is a holistic measure of spatial systems that measures the
ease of reaching destinations {[}10,11{]}.

In practice, the most common form of accessibility measure is based on
the gravity model. These measures are sums of weighted opportunities
around a focal point (i.e., a potential origin), based on how expensive
it is to reach them. Recent research in accessibility analysis has paid
attention to the way opportunities are counted in the pertinent
calculations. Conventionally, the sums are not constrained, which means
that the same opportunity can enter the sum for different origins.
Counting the same opportunity multiple times treats it as if it was
inexhaustible. But opportunities in general are not inexhaustible, and
in fact some of them are by definition exclusive: for example, once a
job is taken up by someone in the population, the same job is no longer
available for any other person to take. More generally, opportunities
are subject to congestion: for example, multiple people can obtain
services from the same family doctor, but the more people who do, the
more congested the service will be.

The issue of congestion in accessibility measures was the motivation for
the development of floating catchment area approaches {[}12,13{]}. While
these approaches purport to account for congestion, Paez et al. {[}14{]}
demonstrate that in general they do not solve the issue of multiple
counting of opportunities, thus leading to biases in the calculation of
total demand and supply, sometimes inflating them, other times deflating
them. In response to this, recent research has paid closer attention to
the way opportunities are counted in accessibility analysis. Paez et al.
{[}14{]}, for example, tackle floating catchment area methods and
introduce a normalization of the impedance matrix to allocate the
population and then the level of service proportionally. More recently,
Soukhov et al. {[}15{]} introduced a singly-constrained measure of
accessibility, called spatial availability, that employs a similar, but
more sophisticated proportional allocation mechanism. The work of these
authors show that floating catchment area methods can be seen as
singly-constrained accessibility measures, and improve on existing
approaches by guaranteeing that each opportunity is counted only once -
in other words, treating opportunities as \emph{finite}. The
proportional allocation of spatial availability constrains the
calculations to match a known quantity, therefore ensuring that the
measurements sum up to a predetermined quantity (i.e., the total number
of opportunities), and so each value can be meaningfully related to this
total.

@@

The objective of this paper is to address this limitation of spatial
availability. Our primary motivation is to extend spatial availability
for the case of multimodal accessibility, but it is worthwhile noting
that this is in fact just one case of heterogeneous populations (i.e.,
travel by different modes). The method itself can easily accommodate
other forms of heterogeneity, for example variations in travel behavior
between older and younger adults {[}e.g., 16{]}, the propensity of older
adults to use different modes of transportation {[}e.g., 17{]}, the
usually shorter trip lengths of children compared to grown-ups {[}e.g.,
18{]}, or the more limited travel ranges of single parents {[}e.g.,
19{]}.

The paper rest of this paper is organized as follows. In Section 2 we
provide a brief review of multimodal accessibility. In the following
Section 3 we demonstrate the derivation of the new spatial availability
expression for multiple modes. In Section 4 we illustrate relevant
issues through a synthetic example. This is followed in Section 5 by an
empirical example using data from the city of Madrid after the
implementation of its Low Emission Zones (LEZ). Data for this example
comes from the city's 2018 travel survey. The example shows the
differences in spatial availability within and outside the LEZ for
travelers using different modes, namely car, transit, cycling and
walking. In Section 6, we provide concluding remarks on the strengths of
the use of spatial availability as a multimodal accessibility measure,
and discuss potential future uses in policy planning scenarios as well
as directions for future research.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used are
measures based on the gravity model {[}20{]}, of which cumulative
opportunity measures and weighted cumulative opportunity measures are
particular forms {[}5{]}. These measures assign a weight to
opportunities based on how easy it is to reach them. Given an origin
(\(i\)) and a destination (\(j\)), an impedance function
\(f^{m}(c^m_{ij})\) converts the cost of travel (e.g., time, money,
generalized cost) into a score that represents the propensity for
interaction. These measures originate from that proposed by Hansen
{[}1{]}, which can take the following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
have mode-specific travel costs (\(c_{ij}^m\)) and/or travel impedance
functions (\(f^m(\cdot)\)).

Hansen-type accessibility is not constrained, which is to say it does
not consider the opportunities as finite. To cite an example, Tahmasbi
et al. {[}21{]} use Hansen-type accessibility to assess the potential
interaction with retail locations by three modes: walking, public
transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of retail
locations \(j\) that can potentially be reached under the travel
impedance as calculated for each \(i\) and \(m\). In other words, for
each origin \(i\) three accessibility scores are calculated. In this
work, Tahmasbi et al. {[}21{]} show that car travel affords the highest
\(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
car can potentially reach more retail opportunities than populations
using other modes. However, higher \(S_i^{m}\) values for car do not
affect the values of \(S_i^{m}\) for other modes: in effect, each mode
is analysed as if the others did not exist. Since the measure is not
constrained, each opportunity is typically counted multiple times within
and between modes, and as a result the sum of accessibility is not
necessarily a meaningful quantity. The accessibility scores for the
modes are often values that are difficult to interpret beyond making
statements about relative size. For example, Lunke {[}22{]}, researching
the region of Oslo, reports accessibility scores for car in the order of
tens of thousands of employment opportunities. The corresponding scores
for transit are lower, but still often in the thousands or tens of
thousands. As reported, the ratio of the transit to the car score can be
lower than 0.2 (meaning transit gives access to less than 20\% of the
opportunities than car). But despite the discussion about ``sufficient
accessibility'', it is unclear what the unconstrained scores mean: is
having access to but 10,000 jobs by transit insufficient? After all,
10,000 employment opportunities are still plenty of opportunities. These
ratios can be found elsewhere in the literature {[}e.g., 23, and
9,16,19, and 8{]}, and they are useful as relative assessment of when
some members of the public are better or worse off than others, but they
do not say much about how bad is ``worse off''.

Besides ratios of accessibility, another way to improve interpretability
of scores sometimes seen in the literature is to standardize them to lie
in the range {[}0-1{]}. This adjustment is only helpful insofar as it
facilitates relative comparisons, but interpretation of the scores
remains challenging because the values are specific to a region and
convey no meaning about the magnitude of the scores. In this approach,
zones always have values between 0 and 1, but how remarkable is a zone
with a low score for pedestrians and a high value for car? And if
remarkable, what does the difference in these standardized values mean
for planners? By how much should transport systems and land-use
configurations be changed to improve conditions? And in what way can
these scores be used to track differences over time? Or between regions?
These questions lack straightforward answers since certain values will
always be relatively `low' or `high', but do not track to a quantity
that can be intuitively understood. Presentation or discussion of
Hansen-type accessibility that has been standardized in this way is not
uncommon in the literature {[}e.g., 24,25{]}.

Once we understand opportunities to be finite, it is possible for an
accessibility measure to take on a crisper meaning. As considered in the
long tradition of accessibility research, capacity of opportunities is
limited and thus is subject to competition by population
{[}12,15,26--30{]}. There are only so many school-seats, hospital
capacity, employment opportunities, etc., in a region and if one person
reaches an opportunity, it is taken: the supply of an opportunity and
the demand for that opportunity are two components of accessibility.
These are clear examples of opportunities that are unambiguously
competitive. But we would go as far as to argue that every type of
opportunity is subject to congestion or capacity constraint, even when
the opportunities are conventionally seen as non-competitive.

Amenities are a good example of this. For instance, standards for
providing green spaces are often stated in the form of \emph{exclusive
access}, in units of amenity per capita. A case in point is a the
Ile-de-France region, a jurisdiction that suggested in a 2013 planning
document that at the municipal level at least \(10m^2\) of public green
space should be supplied \emph{per inhabitant} {[}31{]}. Green spaces
are not evenly distributed, which means that who has access to them
hinges on where they are and how easy is to reach them. This formulation
of provision of amenities is not unusual. For example, Natural England,
an organization that recommends an Accessible Natural Greenspace
Standard such that the minimum supply of space is one ha of statutory
Local Nature Reserves per thousand population\footnote{see
  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
Similarly, the World Health Organization {[}cited in 32{]} recommends
that cities provide a minimum of 9\(m^2\) of green area per inhabitant.
For our purposes, standards of this type translate into ``how much of
this resource is available to one individual that has not been claimed
by anyone else?''. Green spaces often have large capacities, but they
still have a capacity, and it is not the same for a person to have
access to \(5m^2\) of \emph{uncongested} green space as \(15m^2\). This
difference is in fact a matter of justice {[}31,33{]}. Constraining
accessibility is in this way a useful way to evaluate the congested
availability of any type of opportunity. As development of sound
standards is emphasized in the planning literature, in particular in
regards to fairness in transportation {[}see 34{]}, spatial availability
analysis is a useful way to develop and assess standards.

The relevance of the considerations above is put in sharper relief when
we think about the use of multiple modes (or heterogeneous populations).
If we return to Oslo for a moment {[}22{]}, we notice that the places
that have high accessibility by transit are also the places that have
\emph{very high} accessibility by car (in their Figure 2). Those two
populations are going for the same opportunities, and those travelling
by transit have fewer to choose from to begin with. More generally,
people in a zone who are advantaged with relatively low cost of travel
will have the ability to potentially reach more opportunities than other
people. Due to this advantage, through the perspective of finite
opportunities, there are fewer opportunities left for everyone else,
especially for those who use modes that are slower or otherwise more
expensive.

As noted in the Introduction, competitive accessibility was the
rationale for developing floating catchment area methods (FCA),
popularized by Luo et al. {[}13{]} who reformulated the work of Shen
{[}12{]} into two steps (although similar, and earlier, developments are
found in {[}30,35{]}). Shen-type accessibility is formulated as:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
variables are repeated in the Hansen-type measure. Shen-type modal
accessibility (\(a_i^m\)) can be understood as a ratio of the travel
impedance-weighted supply of opportunities for \(m\)-mode in \(i\) over
the travel impedance-weighted demand for opportunities. In this way, it
considers competition. That said, the measure remains unconstrained,
meaning both population \emph{and} opportunities are multiply counted
{[}see 14{]}. In other words, interpretation of the Shen-type
accessibility scores between modes is fraught as it is for Hansen-type
measures.

To illustrate, {[}36{]} calculate \(a_i^m\) to jobs for different
income-group populations in Shenzhen, China using
\(m = \text{public transit}\) and \(m=\text{car}\). Their results
indicate that zones with low-income populations have lower \(a_i^m\)
than zones with higher-income populations. Further, they show that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many zones, arguing that this may further place those zones with
lower-income populations at a disadvantage. \(a_i\) and/or \(a_i^m\) are
used to compare relative spatial differences in overall competitive
accessibility and multimodal competitive accessibility, but because
opportunities were doubly counted (entering the sums of both modes),
this makes for uneasy interpretations of the differences in \(a_i^{m}\)
between modes. Questions that this approach leaves unaddressed include:
what is the impact of competition on the difference in \(a_i^m\) values?
How does the impact vary spatially? And what is the interpretation of
this difference?

Spatial availability improves on previously discussed accessibility
approaches using the Hansen-type measure and the Shen-type measure by
constraining the sum of opportunities, that is, by treating
opportunities as finite. This is done by means of proportional
allocation factors that follow well established principles of spatial
interaction and the gravity model {[}see 37{]}. In Soukhov et al.
{[}15{]} these factors consider: the mass effect (e.g., the size of
populations at different origins); and the cost of travel from different
zones (e.g., some sub-populations face relatively higher or lower
costs). The following section introduces the multimodal form of spatial
availability.

@@

Compared to Hansen-type accessibility:

\begin{equation}
\label{eq:hansen-accessibility}
A_i = \sum_{j=1}^J O_jf(c_{ij})
\end{equation}

\noindent we see that spatial availability is, like the Hansen-type
measure, a weighted sum of the opportunities. What makes spatial
availability stand apart from other approaches is how the weight used in
the sum, balancing factor \(F^t_{ij}\), implements a proportional
allocation mechanisms to ensure that the sum of \(V_i\) is constrained
to match the total number of opportunities in the region. As such,
spatial availability is singly-constrained and naturally implements
competition or congestion. \(F^t_{ij}\) consists of two parts. The first
part is a population-based proportional allocation factor to model the
mass effect of the gravity model: \[
F^p_{i} = \frac{P_i}{\sum_i P_i}
\]

@@

The focus of this paper is to extend \(V_i\) for the measurement of
multimodal applications (or more generally heterogeneous populations).
To do so, the balancing factors need to be reformulated so that 1) the
mass effect now accounts not only for the size of the population at
\(i\), but also the size of sub-populations within \(i\); and 2) the
cost of travel is not only for different zones, but by sub-populations
within each zone (e.g., the cost of travel from \(i\) by car, transit,
walking, etc.) When we introduce modes (or sub-populations) \(m\), the
proportional allocation factors need to satisfy the condition that
\(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be summed across each \(m\) at
each \(i\) and then across all \(i\) to equal to 1. They are also
similarly combined multiplicatively to obtain their joint effect,
represented as the combined balancing factor \(F^{tm}_{ij}\) similar to
that detailed in Equation (\ref{eq:multimodal-balancing-factors}). This
factor is given by:

@@

Next we use a synthetic example to contrast multimodal accessibility and
spatial availability.

@@

Consider the simple system shown in Figure \ref{fig:Fig1}. The figure
shows a region with population at three population centers (\(A\),
\(B\), \(C\)) and jobs at three employment centers (\(1\), \(2\),
\(3\)). The population at each origin \(i\) is consists of two
sub-populations, one using a faster mode \(z\) and another using a
slower mode \(x\), to travel to employment centers. Population center
\(A\) is Suburban: it is closest to its own relatively large employment
center at \(1\), close to the Urban's equally large employment center
\(2\), and has a population that is smaller than the Urban \(B\) and
larger than the Satellite \(C\). \(B\) has the largest \(x\)-using
population, followed by then \(A\), then \(C\). This synthetic example
is inspired by the single-mode example used in {[}12{]} and reconfigured
in {[}15{]}.

@@

From the perspective of access to a \emph{finite} amount of
opportunities in the region (\(210,000\) jobs), the sub-population that
is most proximate to jobs (low cost to reach), furthest from large
populations (less competition), and uses the fastest mode \(z\) (greater
range) can potentially reach the largest number of opportunities. This
appears to be the sub-population at \(A\) using mode \(z\).
Sub-populations located in opposite conditions (i.e., more distant from
jobs, close to large populations, and using slow mode \(x\)) are at a
relative disadvantage. The competition for opportunities between
different mode-using populations matters as it reflects how well the
land-use and transport system serves (or does not serve) certain
populations.

@@

The values calculated for \(S_i^m\) (Hansen-type accessibility),
\(a_i^m\) (Shen-type accessibility), and \(V_i^m\) (spatial
availability) for each \(i\) and \(m\) are shown in the middle three
columns and are aggregated for each \(i\) in the final two columns in
Table 1 . As in the example in Shen {[}12{]}, we use a negative
exponential impedance function
\(f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
\(x\) and \(z\) modes for all accessibility measures calculations.
Notice that in this example we use the same impedance function but the
travel times are different for the two modes. More generally, it is
possible to use different impedance functions for the modes, as
demonstrated in the empirical example in the following section.

Hansen-type accessibility \(S_i^m\) is presented for each origin and
mode in the third column of Table 1 . For all \(i\), the travel by \(z\)
results in higher values of \(S_i^m\) than travel by \(x\). Lack of
competition, or alternatively the assumption of an inexhaustible
resource in the calculation of \(S_i^m\), lead to a curious result.
Since the populations in \(A\) and \(B\) have the same travel impedance
to employment centers \(1\), \(2\) and \(3\) (either 15, 30, or 100
minutes using \(x\) or 10, 25, or 80 minutes using \(z\)), their values
of \(S_i^m\) are the same for both \(A\) and \(B\). Furthermore, the
total sum of \(S_i^m\) in the region is equal to 150,570.2. This value
lacks an intuitive interpretation: it represents the weighted sum of
opportunities that may be reached within the region according to the
travel impedance (i.e., the travel behavior and the characteristics of
the modes) and does not usefully translate into any sort of benchmark.
To connect this example to the aforementioned literature, \(S_i^m\) is
calculated in the work of Tahmasbi et al. {[}21{]}; they contrast
differences in \(S_i^m\) values between modes in a relative and
comparative sense, but make no further interpretation of the \(S_i^m\)
values. More densely populated metropolitan regions will tend to have
more opportunities and hence large \(S_i^m\) values and less densely
regions, smaller values; how much of these differences may simply an
artifact of region density?
